---
title: "Reshaping and Plotting data"
author: "Data Carpentry contributors"
minutes: 20
---

```{r, echo=FALSE}
source("../bin/chunk-options.R")

```

```{r, echo=FALSE, message=FALSE}
library("tidyverse")

gene_name_file <- "https://datadryad.org/bitstream/handle/10255/dryad.85608/scer-mrna-protein-absolute-estimate.txt"

gene_names <- read_tsv(gene_name_file,comment="#") %>%
    select(Gene=gene,Name=orf)

mRNA_data <- read_tsv("ftp://ftp.ncbi.nlm.nih.gov/geo/series/GSE81nnn/GSE81932/suppl/GSE81932_Dataset01.txt.gz")

names(mRNA_data)[1] <- "Name"

mRNA_data <- left_join(mRNA_data,gene_names)
```

------------

> ### Learning Objectives
>
> * Reshape a tibble
> * Start with ggplot2


------------

## Our biological question: can we see the behaviour of our favourite genes?

* Plot the individual trajectories of favourites: NOP16, NOP56, ACT1


## Reshape/tidy the data

```{r, eval = TRUE, purl = TRUE}
gather(mRNA_data,key=Vol,value="log2_ratio",-Name,-Gene)
```

```{r, eval = TRUE, purl = TRUE}
mRNA_data_gathered <- 
gather(mRNA_data,key=Vol,value="log2_ratio",-Name,-Gene) %>%
    separate(Vol,into="Vol_fL") %>%
    mutate(Vol_fL = as.numeric(Vol_fL))

mRNA_data_gathered

```

```{r, eval = TRUE, purl = TRUE}
mRNA_data_NOP16 <- filter(mRNA_data_gathered, Gene=="NOP16")
mRNA_data_NOP16
```

## Plot NOP16 data

```{r, eval = TRUE, purl = TRUE}
ggplot(data=mRNA_data_NOP16, aes(x=Vol_fL,y=log2_ratio)) +
    geom_point()
```


```{r, eval = TRUE, purl = TRUE}
ggplot(data=mRNA_data_NOP16, aes(x=Vol_fL,y=log2_ratio)) +
    geom_line()
```


## Plot all the favourite genes data


```{r, eval = TRUE, purl = TRUE}
mRNA_data_3genes <- filter(mRNA_data_gathered, Gene %in% c("ACT1","NOP16","NOP56"))

ggplot(data=mRNA_data_3genes, aes(x=Vol_fL,y=log2_ratio,colour=Gene)) +
    geom_line()
```

```{r, eval = TRUE, purl = TRUE}
ggplot(data=mRNA_data_3genes, aes(x=Vol_fL,y=Gene,fill=log2_ratio)) +
    geom_tile()
```

```{r, eval = TRUE, purl = TRUE}
ggplot(data=mRNA_data_3genes, aes(x=Vol_fL,y=Gene,fill=log2_ratio)) +
    geom_tile() +
    scale_fill_gradient2()
```

## How to make the plot look nice

```{r, eval = TRUE, purl = TRUE}
ggplot(data=mRNA_data_3genes, aes(x=Vol_fL,y=Gene,fill=log2_ratio)) +
    geom_tile() +
    scale_fill_gradient2() + 
    theme_bw()
```
> ## Challenge: 
>
> 1. Plot your other favourite genes. NOP6?
> 2. Can you change the line colours to something nicer?
>
> > ## Solution
> >
> > ```{r, echo=TRUE}
> > mRNA_data_4genes <- filter(mRNA_data_gathered, Gene %in% c("ACT1","NOP16","NOP56","NOP6"))
> > # ?scale_colour_brewer
> > ```
> {: .solution}
{: .challenge}


{% include links.md %}
