---
title: "Plotting data with ggplot2"
author: "Data Carpentry contributors"
teaching: 60
exercises: 30
objectives:
- "Understand how plots are built by mapping data values to aesthetics"
- "Plot a line plot and heat map from the same data, using ggplot2"

keypoints:
- "Plots are built from mapping data values to position, colour, and other aesthetics."
- "Nice plotting relies on tidy data."
- "The same data can be usefully plotted in different ways."
source: Rmd
---

```{r, echo=FALSE}
source("../bin/chunk-options.R")

```

```{r, echo=FALSE, message=FALSE}
library("tidyverse")

mRNA_data_tidy <- read_csv("../data/mRNA_data_tidy_named.csv")
```

------------

> ### Learning Objectives
>
> * Reshape a tibble
> * Start with ggplot2


------------

## Our biological question: can we see the behaviour of our favourite genes?

* Plot the individual trajectories of favourite genes: NOP16, NOP56, ACT1.
* See how different plotting methods bring out different features of the data.


## Filter only data for NOP16

```{r, eval = TRUE, purl = TRUE}
mRNA_data_NOP16 <- filter(mRNA_data_tidy, Gene=="NOP16")
mRNA_data_NOP16
```

## Plot NOP16 data

Plot only as points.

```{r, eval = TRUE, purl = TRUE}
ggplot(data=mRNA_data_NOP16, aes(x=Vol_fL,y=log2_ratio)) +
    geom_point()
```

Plot as a connected line.

```{r, eval = TRUE, purl = TRUE}
ggplot(data=mRNA_data_NOP16, aes(x=Vol_fL,y=log2_ratio)) +
    geom_line()
```


## Plot all the favourite genes data


```{r, eval = TRUE, purl = TRUE}
mRNA_data_3genes <- filter(mRNA_data_tidy, Gene %in% c("ACT1","NOP16","NOP56"))

ggplot(data=mRNA_data_3genes, aes(x=Vol_fL,y=log2_ratio,colour=Gene)) +
    geom_line()
```

```{r, eval = TRUE, purl = TRUE}
ggplot(data=mRNA_data_3genes, aes(x=Vol_fL,y=Gene,fill=log2_ratio)) +
    geom_tile()
```

```{r, eval = TRUE, purl = TRUE}
ggplot(data=mRNA_data_3genes, aes(x=Vol_fL,y=Gene,fill=log2_ratio)) +
    geom_tile() +
    scale_fill_gradient2()
```

## How to make the plot look nice

```{r, eval = TRUE, purl = TRUE}
ggplot(data=mRNA_data_3genes, aes(x=Vol_fL,y=Gene,fill=log2_ratio)) +
    geom_tile() +
    scale_fill_gradient2() + 
    theme_bw()
```
> ## Challenge: 
>
> 1. Plot your other favourite genes. NOP6?
> 2. Can you change the line colours to something nicer?
>
> > ## Solution
> >
> > ```{r, echo=TRUE}
> > mRNA_data_4genes <- filter(mRNA_data_tidy, Gene %in% c("ACT1","NOP16","NOP56","NOP6"))
> > # ?scale_colour_brewer
> > ```
> {: .solution}
{: .challenge}


{% include links.md %}
